---
title: "Estimating 2020 County Turnout"
author: "Kevin Reuning"
date: "12/11/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
here::here()
```

## Data
The data comes from two places. First, the total votes cast in 2020 by county comes from the [MIT Election Data](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/VOQCHQ). Second, the [Census has estimates](https://www.census.gov/programs-surveys/decennial-census/about/voting-rights/cvap.2019.html) of eligible voting population by county but this is only updated through 2019. The fact is is only updated through 2019 will cause a small problem at the end (fun).

```{r data}
library(data.table)
library(knitr)
vap_dt <- fread("County.csv")
elec_dt <- fread("countypres_2000-2020.csv")

```

### Fix FIPS problems
State FIPS codes are two digit numbers, county FIPS codes are three digit numbers. Often the leading zeros will be lost when CSV or Excel files are used as they are treated as numbers instead of strings. The election dataset has a combined state and county FIPS code where leading zeros are lost. `formatC()` formats a number and returns it as a character. 

The census voting age population data uses a more elaborate `geoid` code where the FIPS code is embedded. `substr()` is used to get just a subset of the string. 
```{r}
elec_dt[,county_fips:=formatC(county_fips, width=5, flag="0")]
head(elec_dt$county_fips)

vap_dt[,county_fips:=substr(geoid, 8, 12)]
head(vap_dt$county_fips)
```

### Selecting just voter turnout data

```{r}
elec_dt_20 <- elec_dt[year ==2020, .(year, state, county_name, county_fips, totalvotes)]
elec_dt_20 <- unique(elec_dt_20)
dim(elec_dt_20)
```

### Selecting just the total VAP data
```{r}
vap_dt <- vap_dt[lntitle=="Total",]
dim(vap_dt)
```

### Check to see what counties are missing
```{r}
missing_fips <- vap_dt[which(!county_fips %in% elec_dt_20$county_fips), county_fips]
table(substr(missing_fips, 0, 2))

missing_fips <- elec_dt_20[which(!county_fips %in% vap_dt$county_fips), county_fips]
table(substr(missing_fips, 0, 2))
```
`02` is Alaska (which is a hint at the problems to come) and `72` is Puerto Rico which should not be in the election data. The other appear to be a handful of random counties that I am guessing are missing because counties change sometimes and FIPS codes change sometimes. 

## Merge data together and calculate turnout 
```{r}
turnout_dt <- merge(vap_dt, elec_dt_20, by="county_fips")
turnout_dt[,turnout_rate:=totalvotes/cvap_est]
range(turnout_dt$turnout_rate, na.rm=T)
turnout_dt[turnout_rate>1 | turnout_rate < 0.1, .(geoname, cvap_est, totalvotes, turnout_rate)]
```

### Fixing Problems
It looks like the MIT election data is providing data using legislative districts but giving it FIPS codes that look like county data, so we will have to remove all the Alaska data. There are about 10 small counties that report higher turnout than voters. This is likely a result of the fact we are using 2019 VAP data instead of 2020. :shrug: 
```{r}
turnout_dt <- turnout_dt[state!="ALASKA",]
```

## Export Data
```{r}
write.csv(turnout_dt[,.(county_fips,county_name, state, turnout_rate, totalvotes, 
              cvap_est, tot_est)], 
          file="county_turnout_2020.csv")
```

You can get it [here](county_turnout_2020.csv)

## Making a map
Legally speaking you cannot combine this data without creating a map even though it has been done a million times before. 

```{r, map}
library(usmap)
library(ggplot2)
names(turnout_dt)[1] <- "fips"
turnout_dt[turnout_rate>1,turnout_rate:=1]
p <- plot_usmap(regions="counties", data=turnout_dt, values="turnout_rate", 
                color = "white", size=.1) 
p + scale_fill_viridis_c("County Turnout", labels=scales::label_percent()) + 
  theme(legend.position = "right")
```

## 2016 Turnout
Lets do it with [2016 now as well](county_turnout_2016.csv). The data comes from the same place but for some reasn the census decided to use all caps for their column labels then. 

```{r}
vap_dt <- fread("County_2016.csv")
vap_dt[,county_fips:=substr(GEOID, 8, 12)]
elec_dt_16 <- elec_dt[year ==2016, .(year, state, county_name, county_fips, totalvotes)]
elec_dt_16 <- unique(elec_dt_16)

vap_dt <- vap_dt[LNTITLE=="Total",]

missing_fips <- vap_dt[which(!county_fips %in% elec_dt_20$county_fips), county_fips]
table(substr(missing_fips, 0, 2))

missing_fips <- elec_dt_20[which(!county_fips %in% vap_dt$county_fips), county_fips]
table(substr(missing_fips, 0, 2))


turnout_dt <- merge(vap_dt, elec_dt_16, by="county_fips")
turnout_dt[,turnout_rate:=totalvotes/CVAP_EST]
range(turnout_dt$turnout_rate, na.rm=T)
turnout_dt[turnout_rate>1 | turnout_rate < 0.1, .(GEONAME, CVAP_EST, totalvotes, turnout_rate)]

turnout_dt <- turnout_dt[state!="ALASKA",]

write.csv(turnout_dt[,.(county_fips,county_name, state, turnout_rate, totalvotes, 
              CVAP_EST, TOT_EST)], 
          file="county_turnout_2016.csv")

```
